---
title: "Checking code style with INWTUtils"
author: "Mira Céline Klein"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking code style}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

This vignette describes how to check your code files for a good style
with `checkStyle` (a wrapper for the `lint` function of the
`lintr` package). The function is tailored to the usage at the INWT Statistics
company but can by applied in other contexts without any disadvantages.

For several so-called *lints* the functions checks if they appear in the code.
In this context, lints are (mostly small) violations of style rules, e.g.,
missing spaces around operators, double spaces, very long lines or trailing
blank lines.
A function checking an specific lint is called linter function.
Section "Included linters" gives more information about the set of tested lints. 


## Why should you watch your style?

Your code may be robust and fast in spite of a bad style. But a good style
makes your code more beautiful and easier to read -- especially for others.
Adapting a consistent style in a team helps to find your way around in the code
written by someone else.

It's never to late to adapt a good coding style -- and never to early.


## How to use `checkStyle`

`checkStyle` can be applied to one or more files.
With the `type` argument you can optionally specify a document type. 
This adds some linters to the set of used linters.
You can choose between scripts (`type = "script"`) or files with package 
functions (`type = "pkgFuns"`). Or you can just ignore the argument.

To demonstrate the usage, we first create two scripts with examples for bad
style:

```{r, echo = FALSE}
library(INWTUtils)
```

```{r badStyleScript, eval = FALSE}
writeLines(c("# This is an example for bad style",
             "x = 1+2",
             "# A comment with  double  spaces",
             "foo <- function( y) print(paste('You entered', y))",
             ""),
           con = "badStyle1.R")
writeLines(c("# This is a second example   ",
             "z<-c(1,2)"),
           con = "badStyle2.R")
```

How many violations of common style conventions do you see? `checkStyle` may
find some more:

```{r checkStyle, results = "hide", eval = FALSE}
checkStyle(files = c("badStyle1.R", "badStyle2.R"),
           type = "script")
```

A new tab opens in RStudio which lists all lints found in the checked files.
It contains the full filepaths and a list with line numbers and
lints for each file. You can start to edit the code and repeat the check
until the opened tab remains empty.

![Output produced by `checkStyle`: For each file you see the full path and a 
list of style rule violations (lints).](style_01checkStyleOutput.PNG)


## Included linters

The following linters are used by default:

```{r lintersAlways, results = 'asis', echo = FALSE}
cat("-", paste0(names(selectLntrs()), collapse = "\n\n- "))
```

If `type = "script"`, the following linters are added:

```{r lintersScript, results = 'asis', echo = FALSE}
cat("-", paste0(names(INWTUtils:::scriptLntrs()), collapse = "\n\n- "))
```

If `type = "pkgFuns"`, the following linters are added:

```{r lintersFuns, results = 'asis', echo = FALSE}
cat("-", paste0(names(INWTUtils:::pkgFunLntrs()), collapse = "\n\n- "))
```

`double_space_linter` checks for double empty spaces.

`internal_INWT_function_linter` checks for the use of internal functions from
packages whose name starts with INWT. Outside of the INWT company, this
linter will barely go into action. [^a]

[^a]: There is usually a reason why an internal function has not been exported.
Either it should not be used in a context outside its package, or the author did
not feel like writing a documentation. To avoid the latter, we want to be
notified about the usage of an internal INWT function so we can add a
documentation to the function and export it.

`setwd_linter` and `source_linter` check for `setwd` or `source` statements
because they can cause side effects when used in functions.

The remaining linters are taken from the `lintr` package.
Details can be found via
[`?lintr::linters`](https://rdrr.io/cran/lintr/man/linters.html) .


## Exclude lines from checking

Sometimes you may want to exclude specific lines from the check because the
found lint cannot be removed for some reason. You achieve this by adding the
`nolint` commands (see also 
[`?lintr::exclude`](https://rdrr.io/cran/lintr/man/exclude.html)):

```{r nolint, eval = FALSE}
# nolint start
x <- c(1,2) # This line will be excluded from the checks
# nolint end
y <- c(3, 4) # This line won't be excluded anymore.
```


```{r rmFiles, echo = FALSE, results = 'hide', eval = FALSE}
lapply(c("badStyle1.R", "badStyle2.R"), unlink)
```
